### 2015-05-21 Harris_2015/manuscript-materials/draft.Rmd
I think this one's actually ready to submit.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500, or
2500 observed communities (24 replicate simulations for each combination; Appendix B).
These simulated data sets span the range from small, single-observer data
sets to large collaborative efforts such as the North American Breeding Bird
Survey. As described in Appendix B, I randomly drew the "true" coefficient
values for each replicate so that most species pairs interacted negligibly, a
few pairs interacted very strongly, and competition was three times more
common than facilitation. I then used Gibbs sampling to randomly generate
landscapes with varying numbers of species and sites via Markov chain Monte
Carlo (Appendix C). For half of the simulated landscapes, I treated each
species' $\alpha$ coefficient as a constant, as described above. For the other
half, I treated the $\alpha$ coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix B). The latter set of simulated landscapes provide an
important test of the methods' ability to distinguish co-occurrence patterns
that were generated from pairwise interactions among the observed species from
those that were generated by external forces like abiotic environmental
filtering. This task was made especially difficult because---as with most
analyses of presence-absence data for co-occurrence patterns---the inference
procedure did not have access to any information about the environmental or
spatial variables that helped shape the landscape [cf @connor_checkered_2013;
@blois_framework_2014].

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
The previous sections involved known values of $\alpha$ and $\beta$. In practice,
ecologists will often need to estimate these parameters from data instead. 
When the number of species is reasonably small, one can compute exact maximum 
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing 
Equation 1 (Appendix C).  Doing so allows us to find the unique set of $\alpha$ and
$\beta$ coefficients that would be expected to produce exactly the observed
co-occurrence frequencies. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. The optimization procedures described here and in Appendix C are 
implemented in the *rosalia* package [@harris_rosalia_2015] for the 
R programming language [@r_core_team_r_2015].

@@

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for nearly a
century, and the models' properties, capabilities, and limits are well-
understood in a huge range of applications, from spatial modeling
[@gelfand_modelling_2005] to neuroscience [@schneidman_weak_2006] to models of
human behavior [@lee_statistical_2013; salakhutdinov_restricted_2007].
Modeling species interactions using the same framework would thus allow
ecologists to tap into an enormous set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states.

>>>>

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500, or
2500 observed communities (24 replicate simulations for each combination; Appendix B).
These simulated data sets span the range from small, single-observer data
sets to large collaborative efforts such as the North American Breeding Bird
Survey. As described in Appendix B, I randomly drew the "true" coefficient
values for each replicate so that most species pairs interacted negligibly, a
few pairs interacted very strongly, and competition was three times more
common than facilitation. I then used Gibbs sampling to randomly generate
landscapes with varying numbers of species and sites via Markov chain Monte
Carlo (Appendix B). For half of the simulated landscapes, I treated each
species' $\alpha$ coefficient as a constant, as described above. For the other
half, I treated the $\alpha$ coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix B). The latter set of simulated landscapes provide an
important test of the methods' ability to distinguish co-occurrence patterns
that were generated from pairwise interactions among the observed species from
those that were generated by external forces like abiotic environmental
filtering. This task was made especially difficult because---as with most
analyses of presence-absence data for co-occurrence patterns---the inference
procedure did not have access to any information about the environmental or
spatial variables that helped shape the landscape [cf @connor_checkered_2013;
@blois_framework_2014].

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
The previous sections involved known values of $\alpha$ and $\beta$. In practice,
ecologists will often need to estimate these parameters from data instead. 
When the number of species is reasonably small, one can compute exact maximum 
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing 
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence 
frequencies. For the analyses in this paper, I used the *rosalia* package 
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to 
define the objective function and gradient as R code.  The rosalia package then 
uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
and $\beta$.

@@

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for nearly a
century, and the models' properties, capabilities, and limits are well-
understood in a huge range of applications, from spatial modeling
[@gelfand_modelling_2005] to neuroscience [@schneidman_weak_2006] to models of
human behavior [@lee_statistical_2013].
Modeling species interactions using the same framework would thus allow
ecologists to tap into an enormous set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states.

### 2015-05-21 Harris_2015/manuscript-materials/draft.tex
I think this one's actually ready to submit.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500,
or 2500 observed communities (24 replicate simulations for each
combination; Appendix B). These simulated data sets span the range from
small, single-observer data sets to large collaborative efforts such as
the North American Breeding Bird Survey. As described in Appendix B, I
randomly drew the ``true'' coefficient values for each replicate so that
most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.
I then used Gibbs sampling to randomly generate landscapes with varying
numbers of species and sites via Markov chain Monte Carlo (Appendix C).
For half of the simulated landscapes, I treated each species' \(\alpha\)
coefficient as a constant, as described above. For the other half, I
treated the \(\alpha\) coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix B). The latter set of simulated landscapes provide
an important test of the methods' ability to distinguish co-occurrence
patterns that were generated from pairwise interactions among the
observed species from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013, Blois et al. 2014).

\textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
presence-absence data.}} The previous sections involved known values of
\(\alpha\) and \(\beta\). In practice, ecologists will often need to
estimate these parameters from data instead. When the number of species
is reasonably small, one can compute exact maximum likelihood estimates
for all of the \(\alpha\) and \(\beta\) coefficients by optimizing
Equation 1 (Appendix C). Doing so allows us to find the unique set of
\(\alpha\) and \(\beta\) coefficients that would be expected to produce
exactly the observed co-occurrence frequencies. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
(Murphy 2012), ensuring that this procedure will always converge on the
global maximum. The optimization procedures described here and in
Appendix C are implemented in the \emph{rosalia} package
({\textbf{???}}) for the R programming language (R Core Team 2015).

@@

Apart from the environmental filters, the simulated landscapes presented
here represent the best-case scenario for these methods. Future research
should thus examine these models' performance characteristics when the
``true'' interaction matrices include guild structure or trophic levels,
which could make the \(\beta\) coefficients much more difficult to infer
(particularly for linear approximations like the partial covariance
approach; Loh and Wainwright (2013)). On the other hand, ecologists may
often have prior information about the nature of real species'
interaction patterns from natural history or ecological experiments,
which could substantially reduce the probability and magnitude of error.
The rosalia package ({\textbf{???}}) has built-in mechanisms for
incorporating this kind of information, if it can be expressed as a
prior probability distribution or a penalty on the likelihood.

@@

Despite these limitations, the results with environmental filtering seem
to indicate that the method can be very robust. Additionally, the fact
that Markov networks provide a likelihood function to optimize makes
them highly extensible, even when it is inconvenient to compute the
likelihood exactly. For example, the mistnet software package for joint
species distribution modeling (Harris 2015) can fit \emph{approximate}
Markov networks to large species assemblages (\textgreater{}300 species)
while simultaneously modeling each species' response to the abiotic
environment with complex, nonlinear functions. This sort of approach,
which combines multiple ecological processes, could help ecologists to
disentangle different factors behind the co-occurrence patterns we
observe in nature. Numerous other extensions are possible: similar
networks can be fit with continuous variables, count data, or both (Lee
and Hastie 2012). There are even methods (Whittam and Siegel-Causey
1981, Tjelmeland and Besag 1998) that would allow the coefficient
linking two species in an interaction matrix to vary as a function of
the abiotic environment or of third-party species that could tip the
balance between facilitation and exploitation (Bruno et al. 2003). Fully
exploring these possibilities will require more research into the
various available approximations to the log-likelihood and to its
gradient, in order to balance efficiency, accuracy, and the ability to
generate confidence limits for statistical inference.

@@

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for
nearly a century, and the models' properties, capabilities, and limits
are well- understood in a huge range of applications, from spatial
modeling (Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006)
to models of human behavior {[}Lee et al. (2013);
salakhutdinov\_restricted\_2007{]}. Modeling species interactions using
the same framework would thus allow ecologists to tap into an enormous
set of existing discoveries and techniques for dealing with indirect
effects, stability, and alternative stable states.

@@

Harris, D. J. 2015. Generating realistic assemblages with a Joint
Species Distribution Model. Methods in Ecology and Evolution.

@@

Lewin, R. 1983. Santa Rosalia Was a Goat Ecologists have for two decades
made assumptions about the importance of competition in community
organization; that idea is now under vigorous attack. Science
221:636--639.

@@

Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
\end{flushleft}
\end{spacing}

>>>>

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500,
or 2500 observed communities (24 replicate simulations for each
combination; Appendix B). These simulated data sets span the range from
small, single-observer data sets to large collaborative efforts such as
the North American Breeding Bird Survey. As described in Appendix B, I
randomly drew the ``true'' coefficient values for each replicate so that
most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.
I then used Gibbs sampling to randomly generate landscapes with varying
numbers of species and sites via Markov chain Monte Carlo (Appendix B).
For half of the simulated landscapes, I treated each species' \(\alpha\)
coefficient as a constant, as described above. For the other half, I
treated the \(\alpha\) coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix B). The latter set of simulated landscapes provide
an important test of the methods' ability to distinguish co-occurrence
patterns that were generated from pairwise interactions among the
observed species from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013, Blois et al. 2014).

\textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
presence-absence data.}} The previous sections involved known values of
\(\alpha\) and \(\beta\). In practice, ecologists will often need to
estimate these parameters from data instead. When the number of species
is reasonably small, one can compute exact maximum likelihood estimates
for all of the \(\alpha\) and \(\beta\) coefficients by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will always converge on the global maximum.
This maximum is the unique combination of \(\alpha\) and \(\beta\)
coefficients that would be expected to produce exactly the observed
co-occurrence frequencies. For the analyses in this paper, I used the
\emph{rosalia} package (Harris 2015a) for the R programming language (R
Core Team 2015) to define the objective function and gradient as R code.
The rosalia package then uses the \texttt{BFGS} method in R's
\texttt{optim} function to find the best values for \(\alpha\) and
\(\beta\).

@@

Apart from the environmental filters, the simulated landscapes presented
here represent the best-case scenario for these methods. Future research
should thus examine these models' performance characteristics when the
``true'' interaction matrices include guild structure or trophic levels,
which could make the \(\beta\) coefficients much more difficult to infer
(particularly for linear approximations like the partial covariance
approach; Loh and Wainwright (2013)). On the other hand, ecologists may
often have prior information about the nature of real species'
interaction patterns from natural history or ecological experiments,
which could substantially reduce the probability and magnitude of error.
The rosalia package (Harris 2015a) has built-in mechanisms for
incorporating this kind of information, if it can be expressed as a
prior probability distribution or a penalty on the likelihood.

@@

Despite these limitations, the results with environmental filtering seem
to indicate that the method can be very robust. Additionally, the fact
that Markov networks provide a likelihood function to optimize makes
them highly extensible, even when it is inconvenient to compute the
likelihood exactly. For example, the mistnet software package for joint
species distribution modeling (Harris 2015b) can fit \emph{approximate}
Markov networks to large species assemblages (\textgreater{}300 species)
while simultaneously modeling each species' response to the abiotic
environment with complex, nonlinear functions. This sort of approach,
which combines multiple ecological processes, could help ecologists to
disentangle different factors behind the co-occurrence patterns we
observe in nature. Numerous other extensions are possible: similar
networks can be fit with continuous variables, count data, or both (Lee
and Hastie 2012). There are even methods (Whittam and Siegel-Causey
1981, Tjelmeland and Besag 1998) that would allow the coefficient
linking two species in an interaction matrix to vary as a function of
the abiotic environment or of third-party species that could tip the
balance between facilitation and exploitation (Bruno et al. 2003). Fully
exploring these possibilities will require more research into the
various available approximations to the log-likelihood and to its
gradient, in order to balance efficiency, accuracy, and the ability to
generate confidence limits for statistical inference.

@@

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for
nearly a century, and the models' properties, capabilities, and limits
are well- understood in a huge range of applications, from spatial
modeling (Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006)
to models of human behavior (Lee et al. 2013). Modeling species
interactions using the same framework would thus allow ecologists to tap
into an enormous set of existing discoveries and techniques for dealing
with indirect effects, stability, and alternative stable states.

@@

Harris, D. J. 2015a. Rosalia: Exact inference for small binary Markov
networks. R package version 0.1.0. Zenodo.
http://dx.doi.org/10.5281/zenodo.17808.

Harris, D. J. 2015b. Generating realistic assemblages with a Joint
Species Distribution Model. Methods in Ecology and Evolution.

@@

Lewin, R. 1983. Santa Rosalia Was a Goat. Science 221:636--639.

@@

Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.

\end{flushleft}
\end{spacing}

### 2015-05-21 Harris_2015/models.Rmd
version bump, manuscript tweaks, final figure
I think this one's actually ready to submit.

---
output: html_document
---
# Simulating landscapes from known coefficients

@@

@@

# Model fitting

@@

## R-compatible methods

@@

### Rosalia

@@

### Partial covariance

@@

### Sample covariance

@@

## Analyzing simulated landscapes with the first three models

>>>>

---
output: html_document
---
# Analyses for Figure 3

## Simulating landscapes from known coefficients

@@

I used Gibbs sampling as my Markov chain Monte Carlo transition operator.
In each round of Gibbs sampling, I cycled through all the species, randomly
updating each one's presence/absence vector in response to its conditional
occurrence probability:

$$p(y_i) = \mathrm{logistic}(\alpha_i + \sum_j\beta_{ij}y_j),$$

where the logistic function is $\frac{1}{1+e^{-x}}$.

@@

## Model fitting

@@

### R-compatible methods

@@

***Rosalia***

@@

***Partial covariance***

@@

***Sample covariance***

@@

### Analyzing simulated landscapes with the first three models
